---
title: "Homework 3"
author: "Sascha Roberts"
format: pdf
---

# 1. Set up

```{r}
#| label: libraries-and-data
#| message: false

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(flextable) # creating tables

# create kelp object with data
kelp <- read_csv(here("data", "temp-kelp.csv"))
# create bike_commute data with personal data
bike_commute <- read_xlsx(here("data", "ENVS_193DS_Personal_Data_Project.xlsx"))
```


# 2. Problems

## Problem 1. Giant kelp fronds

### a. An appropriate test

Because we are interested in the strength of the relationship between temperature and giant kelp
frond elongation rate, the appropriate test is a Pearson’s r or a Spearman rank correlation. These tests both test for correlation between two variables and see haw the variables move together, not necessarily predict each other. Pearson’s r test is a parametric test that can be used for linear relationships that have continuous and normally distributed variables. Spearman rank correlation is a non-parametric test, which can be used for monotonic relationships that have interval, ordinal, or nominal data.

### b. Create a visualization

```{r}
#| label: kelp_visualization

# base layer: ggplot
# using kelp data frame
ggplot(data = kelp, 
       mapping = aes(
          x = temp_c, # x- axis: Ocean temperature (degrees C)
          y = kelp_elong)) + # y axis: kelp elongation rate (cm/day)
  # first layer: points
  geom_point(
      #changing color from default
      color = "darkcyan") + 
  # relabeling x- and y- axes
  labs(
      x = "Ocean Temperature (°C)", #
      y = "Kelp Frond Elongation Rate (cm/day)") +
  # changing theme from default
  theme_bw()

```

### c. Check your assumptions and run your test

#### Linearity

```{r}
#| label: kelp_linearity

# base layer: ggplot
# using kelp data frame
ggplot(data = kelp, 
       mapping = aes(
          x = temp_c, # x- axis: Ocean temperature (degrees C)
          y = kelp_elong)) + # y axis: kelp elongation rate (cm/day)
  # first layer: points
  geom_point(
      #changing color from default
      color = "darkcyan") + 
  # relabeling x- and y- axes
  labs(
      x = "Ocean Temperature (°C)", #
      y = "Kelp Frond Elongation Rate (cm/day)") +
  # changing theme from default
  theme_bw()

```

#### Model fitting

```{r}
#| label: kelp-model

kelp_model <- lm(
  kelp_elong ~ temp_c, # formula: change in kelp length as a function of temp
  data = kelp)      # data: kelp data

```

#### Diagnostics

```{r}
#| label: kelp_diagnostics

par(mfrow = c(2, 2))   # displays plots in a 2 by 2 grid
plot(kelp_model)    # displays diagnostic plots for model

```

The data is linear and continuous meaning a pearsons r test should be used. In the diagnostic plots, the Q-Q Residuals plot shows that the residuals are normally distributed. The Residuals vs Fitted and Scale-Location plots show that the residuals are evenly distributed with no clear patterns. And finally, the Residuals vs Leverage plot shows that their are no outliers in the data that may have a large influence on the models. 

#### Running test
```{r}
#| label: kelp_test

# running correlation test
# variables ocean temperature and giant kelp elongation rate
cor.test(kelp$temp_c, kelp$kelp_elong, 
    # running pearson's r based on assumptions
    method = "pearson") 

```
 
### d. Results communication

To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used a

To evaluate the strength of the relationship between ocean temperature and giant kelp (Macrocystis pyrifera) frond elongation rate, I used a Pearson’s r correlation test. The test revealed a moderate to strong negative (Pearson’s r = -0.69) relationship between ocean temperature and kelp elongation rate(cm/day) (t(30) = -5.19, p < 0.001,  $\alpha$ = 0.05).

### e. Test implications

Our results revealed a strong to moderate negative relationship between ocean temperature and Giant Kelp Elongation rate. This means that at higher ocean temperatures, Giant Kelp frond elongation rate is lower and therefore, to maximize Giant Kelp growth, ocean water temperature should be at a natural low. As average ocean temperature rises due to climate change, we will most likely see a decrease in Giant Kelp Elongation rate.




