---
title: "Homework 3"
author: "Sascha Roberts"
format: pdf
---

# 1. Set up

```{r}
#| label: libraries-and-data

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(flextable) # creating tables

# create kelp object with data
kelp <- read_csv(here("data", "temp-kelp.csv"))
# create bike_commute data with personal data
bike_commute <- read_xlsx(here("data", "ENVS_193DS_Personal_Data_Project.xlsx"))
```


# 2. Problems

## Problem 1. Giant kelp fronds

### a. An appropriate test

Because we are interested in the strength of the relationship between temperature and giant kelp
frond elongation rate, the appropriate test is a Pearson’s r or a Spearman rank correlation. These tests both test for correlation between two variables and see haw the variables move together, not necessarily predict each other. Pearson’s r test is a parametric test that can be used for continuous and normally distributed variables. Spearman rank correlation is a non-parametric test, which can be used for interval, ordinal, or nominal data.

### b. Create a visualization

```{r}
#| label: kelp_visualization

# base layer: ggplot
# using kelp data frame
ggplot(data = kelp, 
       mapping = aes(
          x = temp_c, # x- axis: Ocean temperature (degrees C)
          y = kelp_elong)) + # y axis: kelp elongation rate (cm/day)
  # first layer: points
  geom_point(
      #changing color from default
      color = "darkcyan") + 
  # relabeling x- and y- axes
  labs(
      x = "Ocean Temperature (°C)", #
      y = "Kelp Frond Elongation Rate (cm/day)") +
  # changing theme from default
  theme_bw()

```

### c. 
